---
title: 'Reproducible Research: Peer Assessment 1'
author: "Aldo Parra"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = F)
```

## Load packages


```{r}
library(lubridate)
library(dplyr)
library(ggplot2)
```


## Loading and preprocessing the data

Unzip the data and load the csv file in a new object, `activity`. Set the `na.strings` argument equal to `"NA"` due missing values are coded as *NA* 

```{r}
unzip("activity.zip")
activity <- read.csv("activity.csv", stringsAsFactors = F, na.strings = "NA")
head(activity)
```

Change the **date** column class to Date format

```{r}
activity$date <- ymd(activity$date)
```

Delete missing values and save a data set with missing values for future analysis

```{r}
activity_na <- activity
activity <- activity[complete.cases(activity),]
```

## What is mean total number of steps taken per day?

dplyr functions are used to get the total number of steps taken every day

```{r}
steps_sum <- activity %>% group_by(date) %>% 
    summarise(Total_steps = sum(steps)) %>% 
    print()
```


Mean and median 
```{r}
mean(steps_sum$Total_steps)
median(steps_sum$Total_steps)
```

```{r}
ggplot(steps_sum, aes(Total_steps)) +
    geom_histogram() +
    labs(title = "Total number of steps taken per day histogram", x = "Number of steps", y = "Freq")
```


## What is the average daily activity pattern?

Number of average steps per day
```{r}
steps_m <- activity %>% group_by(interval) %>% 
    summarise(Steps_mean = mean(steps))
```


Time series plot
```{r}
ggplot(steps_m, aes(interval, Steps_mean)) +
    geom_line(size = 1) +
    labs(title = "Daily activity pattern", y = "Average number of steps", x = "Interval")
```


## Imputing missing values

Total number of missing values in the dataset
```{r}
sum(!complete.cases(activity_na))
```

Filling in all of the missing values in the dataset with the mean of steps
```{r}
activity_na$steps[is.na(activity_na$steps)] <- mean(activity$steps)
```


New dataset equal to the original dataset with the missing data filled in.
```{r}
activity_full <- activity_na 
```

Histogram, mean and median of the total number of steps taken each day 

```{r}
steps_sum_full <- activity_full %>% group_by(date) %>% 
    summarise(Total_steps = sum(steps))

mean(steps_sum_full$Total_steps)
median(steps_sum_full$Total_steps)

ggplot(steps_sum_full, aes(Total_steps)) +
    geom_histogram() +
    labs(title = "Total number of steps taken per day histogram", x = "Number of steps", y = "Freq")
```


Notice that there is not significant difference with the previous results due the missing data imputation. This fact is a consequence of the use of the global mean to replace missing data


```{r}
mean(steps_sum_full$Total_steps) - mean(steps_sum$Total_steps)
median(steps_sum_full$Total_steps) - median(steps_sum$Total_steps)
```

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend"
```{r}
activity_full$Weekend <- ifelse(weekdays(activity_full$date) == "sábado" | weekdays(activity_full$date) == "domingo", "Weekend", "Weekday")
activity_full$Weekend <- as.factor(activity_full$Weekend)
```

Panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
steps_mw <- activity_full %>% group_by(interval, Weekend) %>% 
    summarise(Steps_mean = mean(steps))

ggplot(steps_mw, aes(interval, Steps_mean)) +
    geom_line(size = 1) +
    facet_wrap(~Weekend, ncol = 1) +
    labs(title = "Average number of steps taken per interval", y = "Average number of steps", x = "Interval")
```

The previus plot shows that there are differenciences in activity patterns between weekdays and weekends. There is less activity at weekend between 0 and 1000 intervals, however, the activity at weekend is greater than on a weekday between the 1000 and highger intervals

Info about the session
```{r}
sessionInfo()
```
